---
title: "Analysis of Categorical Data (cont.)"
author: "NRES 710"
date: "Last compiled: `r Sys.Date()`"
output: 
  html_document: 
    theme: yeti
    toc: yes
    toc_float: yes
    css: styles.css
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE
)
```

```{r echo=FALSE}

#  NRES 710, Analysis of Categorical Data (cont.)
#     University of Nevada, Reno
#     Presenting results and multiple groups

```

## Summary

Last class we discussed how to analyze data when we have a categorical X-variable data. 

Traditionally, this is called a *t-test* -- but I am going to avoid calling it that going forward. Instead, we are always going to be using the **linear model** in this class, which is more *parsimonious* (simple) and will create continuity with the approach we are going to take toward understanding statistics in this class 

Today we will discuss how to **present results** from analysis of categorical data. We have a 'canned' sentence for presenting results when we have two continuous variables, but we need a slightly different approach for when we have categorical X-data.

The good news is that our X-variable data can only be continuous or categorical. So once we learn this second approach, we will only have to know these **two approaches**. We can use them repeatedly during our various analytical activities, perhaps with slight modifications.

## Presenting Results

With continuous X data, we start out by saying: "For each 1 unit increase in X, we observed a unit change in Y." There is no such thing as a unit change in X when we have a categorical X variable, so we will drop that. Instead, we will say something like:

**"We found that [group 1] was [$\beta_1$] [Y-units] ([+/-CI]; 95% CI) [descriptor word to indicate direction] than [group 2] (p = [p-value])."

- *if p > 0.05, then add: "; however, our results were not statistically significant."

### Example

Let's take a peak in R at the elephant [seal body size data](lecture_8_seal_data.csv) from last class:

```{r seals, echo=TRUE, message=FALSE, warning=FALSE, fig.width=4.5, fig.height=3.5}
# Load the data
datum <- read.csv("lecture_8_seal_data.csv")

# Examine it
head(datum)
tail(datum)
```






[--go to next lecture--](lecture_10.html)
