---
title: "Schedule"
author: "NRES 710"
date: "Fall 2024"
output: 
  html_document: 
    theme: yeti
    toc: no
    toc_float: no
---

<style type="text/css">
body{ /* Normal  */ font-size: 16px; font-family: Arial; color:black}
th { /* Table Header */ font-size: 16px; /}
td {  /* Table  */ font-size: 14px;}
h1.title {font-size: 30px; color: Black; font-weight: bold;}
.author {font-size: 30px; color: Black; font-weight: bold;}
h1 { /* Header 1 */ font-size: 28px; color: Black; font-weight: bold;}
h2 { /* Header 2 */ font-size: 22px; color: Black; font-weight: bold;}
h3 { /* Header 3 */ font-size: 18px; color: Black; font-weight: bold;}
h4 { /* Header 4 */ font-size: 16px; color: Black; font-weight: bold;}
code.r{ /* Code block */ font-size: 12px;}
pre { /* Code block - determines code spacing between lines */ font-size: 14px;}
.main-container {padding-left: 25px;}
.toc-content {
  width: 75%; /* Increase the width of the main content area */
  margin-left: 27%; /* Ensure this aligns properly with the TOC width */
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Schedule

Note: this schedule is subject to change. Please check for updates frequently. Also, if a due date listed here differs from that on WebCampus, please use the date listed on WebCampus! Thank you.

```{r include=FALSE}
courseSchedule <- read.csv("CourseSchedule.csv")

```

```{r results='asis', echo=FALSE}
knitr::kable(courseSchedule, caption="", col.names = c("Week", "Tuesday",
                                                     "Thursday (R mini-lab)",
                                                     "Assignments"))
```

